The Perils of Exploration under Competition: \\ A Computational Modeling Approach

### 2020-07-15 journal paper/content/sec-bg.tex
Minor edits.

We present sufficient background on the three classes of MAB algorithms listed in Section~\ref{sec:intro-model}: from adaptive-exploration to exploration-separating to greedy.

@@

A standard performance measure is \emph{regret}, defined as the difference in the total expected reward between the algorithm and the best arm.%
\footnote{In a formula, regret is
    $T\cdot \max_{\text{arms $a$}}\mu_a
    -  \E\sbr{ \sum_{t\in[T]} r_t }$,
where $\mu_a$ is the mean reward of arm $a$.}
Since regret is normalized by the best arm, it allows to compare algorithms across different problem instances.
The primary concern is the asymptotic growth rate of regret as a function of the time horizon $T$.

The three classes of algorithms perform very differently in terms of regret. Adaptive-exploration algorithms achieve optimal regret rates:
    $\tildeO(\sqrt{KT})$
for all problem instances, and an improved regret rate of
    $O(\tfrac{K}{\Delta}\log T)$
for problem instances with $\text{gap}\geq \Delta$
\citep{Lai-Robbins-85,bandits-ucb1,bandits-exp3}.%
\footnote{The \emph{gap} is the difference in mean reward between the best arm and the second-best arm. As usual in computer science, the $O(\cdot)$ and $\tildeO(\cdot)$ notation suppresses the constant and
$\polylog(T)$ factors, respectively.}
Exploration-separating algorithms can only achieve regret $\tildeO(T^{2/3})$ across all problem instances. Finally, the greedy algorithm is terrible on a wide variety of problem instances, in the sense that it fails to try the best arm and therefore suffers regret linear in $T$
\citep[see Chapter 11.2 in][]{slivkins-MABbook}.

@@

Exploration-separating algorithms completely separate exploration and exploitation. The simplest approach, called \emph{Explore-First}, is to explore uniformly for a predetermined number of rounds, then choose one arm and ``exploit". A slightly more refined approach, called \emph{Epsilon-Greedy}, explores uniformly in each round $t$ with predetermined probability $\eps_t$, and exploits with the remaining probability. A formal definition is that each round is either selected exploration, in which case  the distribution over arms does not depend on data, or it is assigned to exploitation, in which case the data from this round is discarded. The $T^{2/3}$ lower bound on regret follows from a somewhat more general lower bound in \citet{MechMAB-ec09}.

\xhdr{Advanced aspects.}
Switching from ``greedy" to ``exploration-separating" to ``adaptive-exploration" algorithms involves substantial adoption costs in infrastructure and personnel training \citep{DS-arxiv}. Inserting exploration in a complex decision pipeline necessitates a substantial awareness of the technology and a certain change in a mindset, as well as an infrastructure to collect and analyze the data. Adaptive exploration requires the said infrastructure to propagate the data analysis back to the ``front-end" where the decisions are made, and do it on a sufficiently fast and regular cadence. Framing the problem in terms of the appropriate modeling assumptions and action features, and debugging the machine learning code, tend to be quite subtle, too.

@@

\xhdr{Instantaneous regret.}

@@

>>>>

We present self-contained background on multi-armed bandits (MAB), to make the paper accessible to non-specialists. We focus on the three classes of MAB algorithms listed in Section~\ref{sec:intro-model}: from adaptive-exploration to exploration-separating to greedy.

@@

A standard performance measure is \emph{regret}, defined as the difference in the total expected reward between the algorithm and the best arm. In a formula, regret is
    $T\cdot \max_{\text{arms $a$}}\mu_a
    -  \E\sbr{ \sum_{t\in[T]} r_t }$,
where $\mu_a$ is the mean reward of arm $a$.
Normalized by the best arm, regret allows to compare algorithms across different problem instances.
The primary concern is the asymptotic growth rate of regret as a function of $T$.

The three classes of algorithms perform very differently in terms of regret: adaptive-exploration algorithms are by far the best, greedy algorithms are by far the worst, and exploration-separating ones are in the middle. The precise regret rates are not essential to this paper; we present them for illustration only. Adaptive-exploration algorithms achieve optimal regret rates:
    $\tildeO(\sqrt{KT})$
for all problem instances, and an improved regret rate of
    $O(\tfrac{K}{\Delta}\log T)$
for problem instances with $\text{gap}\geq \Delta$
\citep{Lai-Robbins-85,bandits-ucb1,bandits-exp3}.%
\footnote{The \emph{gap} is the difference in mean reward between the best arm and the second-best arm. As usual in computer science, the $O(\cdot)$ and $\tildeO(\cdot)$ notation suppresses the constant and
$\polylog(T)$ factors, respectively.}
Exploration-separating algorithms can only achieve regret $\tildeO(T^{2/3})$ across all problem instances. Finally, the greedy algorithm is terrible on a wide variety of problem instances, in the sense that it fails to try the best arm and therefore suffers regret linear in $T$
\citep[see Chapter 11.2 in][]{slivkins-MABbook}.

@@

Exploration-separating algorithms completely separate exploration and exploitation. Each round is either selected for exploration, in which case  the distribution over arms does not depend on data, or it is assigned to exploitation, in which case the data from this round is discarded. The simplest approach, called \emph{Explore-First}, is to explore uniformly for a predetermined number of rounds, then choose one arm and ``exploit". A slightly more refined approach, called \emph{Epsilon-Greedy}, explores uniformly in each round with a predetermined probability, and ``exploits" with the remaining probability. The $T^{2/3}$ lower bound on regret follows from a somewhat more general lower bound in \citet{MechMAB-ec09}.

\xhdr{Some advanced aspects.}
Switching from ``greedy" to ``exploration-separating" to ``adaptive-exploration" algorithms involves substantial adoption costs in infrastructure and personnel training \citep{DS-arxiv}. Inserting exploration in a complex decision pipeline necessitates a substantial awareness of the technology and a certain change in a mindset, as well as an infrastructure to collect and analyze the data. Adaptive exploration requires the said infrastructure to propagate the data analysis back to the ``front-end" where the decisions are made, and do it on a sufficiently fast and regular cadence. Framing the problem in terms of the appropriate modeling assumptions and action features, and debugging the machine learning code, tend to be quite subtle, too.

@@

\xhdr{Instantaneous regret.}
Cumulative performance measures such as regret are not quite appropriate for our setting, as we need to characterize interactions in particular rounds. Instead, our theoretical results focus on \emph{Bayesian instantaneous regret} (\BIR), as defined in Section~\ref{sec:theory-prelims}. Recall that we posit a Bayesian prior on the \MRVs. In the notation of this appendix, the \BIR is simply:
\begin{align*}%\label{eq:bg-BIR-defn}
\BIR(t) := \E_{\text{prior}}\sbr{ \max_{a\in A} \mu_a - r_t}.
\end{align*}
\noindent Note that Bayesian regret (\ie regret in expectation over the prior) is precisely
\begin{align}\label{eq:bg-BReg}
  \BReg(T):= 
    \E_{\text{prior}}\sbr{ T\cdot \max_{\text{arms $a$}}\mu_a - \sum_{t=1}^T r_t }
    = \sum_{t=1}^T \BIR(t).
\end{align}
We are primarily interested in how fast \BIR decreases with $t$.%
\footnote{We treat $K$, the number of arms, as a constant in what follows.}

The three classes of algorithms are well-separated in terms of \BIR, much like they are in terms of regret.

\begin{itemize}
\item \DynGreedy has at least a constant \BIR for many reasonable priors (where the constant can depend on $K$ and the prior, but not on $t$). The reason and the proof are exactly the same as for regret.

\item Exploration-separating algorithms can achieve
    $\BIR(t) = \tildeO(t^{-1/3})$
for all priors, \eg by using Epsilon-Greedy with exploration probability $\eps_t = t^{-1/3}$ in each round $t$. 
The $t^{-1/3}$ rate is the best possible, in the sense that

@@

\item Adaptive-exploration algorithms can have an even better rate
    $\BIR(n) \sim n^{-1/3}$
for all priors. 
\end{itemize}

Both rates are up to $\log(n)$ factors, and are the best possible \BIR rates that can be achieved for all priors.
\ascomment{stopped here.}

### 2020-07-15 journal paper/content/sim_details.tex
Minor edits.

\xhdr{MAB algorithms.} We consider the three classes of bandit algorithms discussed in Section~\ref{sec:intro-model}, and look for qualitative differences between them. We take a representative algorithm from each class. Our pilot experiments indicate that our findings do not change substantially if other representative algorithms are chosen. 
We use \Thompson (\TS) from the ``adaptive-exploration" algorithms, \DynamicEpsGreedy (\DEG) with exploration probability $\eps=5\%$ from the ``exploration-separating" algorithms,%
\footnote{In each round, \DynamicEpsGreedy algorithm explores uniformly with a predetermined probability $\eps$, and ``exploits" with the remaining probability, choosing an arm with maximal posterior mean reward given the current data. Our pilot experiments show that choosing a different exploration probability $\eps$ does not qualitatively change the results.}  and \DynGreedy (\DG) algorithm  as in Section~\ref{sec:theory-prelims}. For ease of comparison, all three algorithms are parameterized with the same ``fake" Bayesian prior: namely, the mean reward of each arm is drawn independently from a $\Beta(1,1)$ distribution. Recall that Beta priors with 0-1 rewards form a conjugate family, which allows for simple posterior updates.
Self-contained background can be found in Appendix~\ref{sec:related-classes}.

>>>>

\xhdr{MAB algorithms.} We consider the three classes of bandit algorithms discussed in Section~\ref{sec:intro-model}, and look for qualitative differences between them. We take a representative algorithm from each class. Our pilot experiments indicate that our findings do not change substantially if other representative algorithms are chosen.
We use \Thompson (\TS) from the ``adaptive-exploration" algorithms, \DynamicEpsGreedy (\DEG) from the ``exploration-separating" algorithms,%
\footnote{In each round, \DynamicEpsGreedy algorithm explores uniformly with a predetermined probability $\eps$, and ``exploits" with the remaining probability, choosing an arm with maximal posterior mean reward given the current data. We use $\eps=5\%$ throughout. Our pilot experiments show that choosing a different $\eps$ does not qualitatively change the results.}  and \DynGreedy (\DG) algorithm  as in Section~\ref{sec:theory-prelims}. For ease of comparison, all three algorithms are parameterized with the same ``fake" Bayesian prior: namely, the mean reward of each arm is drawn independently from a $\Beta(1,1)$ distribution. Recall that Beta priors with 0-1 rewards form a conjugate family, which allows for simple posterior updates.
Self-contained background can be found in Appendix~\ref{sec:related-classes}.

### 2020-07-15 journal paper/ec19paper/content/inverted_u.tex
Minor edits.

\xhdr{Inverted-U relationship.}
We interpret our findings through the lens of the inverted-U relationship between the ``intensity of competition" and the ``quality of technology". The lowest level of competition is monopoly, when \DynamicGreedy wins out for the trivial reason of tie-breaking. The highest levels are simultaneous entry and ``late entrant". We see that \DynamicGreedy is incentivized for exploration-disadvantaged MAB instances. In fact, incentives for \DynamicGreedy get stronger when the model transitions from simultaneous entry to ``late entrant".%
\footnote{For the Heavy-Tail instance, \DynamicGreedy goes from a weakly dominant strategy to a strictly dominant one. For the Uniform instance, \DynamicGreedy goes from a Nash equilibrium strategy to a weakly dominant one.}
Finally, the middle level of competition, ``incumbent" in the first-mover regime creates strong incentives for \Thompson. In stylized form, this relationship is captured in Figure~\ref{fig:inverted-U-expts}.%
\footnote{\asedit{We consider the monopoly scenario for comparison only, without presenting any findings. We just assume that a monopolist chooses
the greedy algorithm, because it is easier to deploy in practice. Implicitly, users have no ``outside option": the service provided is an improvement over not having it (and therefore the monopolist is not incentivized to deploy better learning algorithms). This is plausible with free ad-supported platforms such as Yelp or Google.}}

>>>>

\xhdr{Inverted-U relationship.}
We interpret our findings through the lens of the inverted-U relationship between the ``intensity of competition" and the ``quality of technology". The lowest level of competition is monopoly, when \DynamicGreedy wins out for the trivial reason of tie-breaking. The highest levels are simultaneous entry and ``late entrant". We see that \DynamicGreedy is incentivized for exploration-disadvantaged MAB instances. In fact, incentives for \DynamicGreedy get stronger when the model transitions from simultaneous entry to ``late entrant".%
\footnote{For the Heavy-Tail instance, \DynamicGreedy goes from a weakly dominant strategy to a strictly dominant one. For the Uniform instance, \DynamicGreedy goes from a Nash equilibrium strategy to a weakly dominant one.}
Finally, the middle level of competition, ``incumbent" in the first-mover regime creates strong incentives for \Thompson. In stylized form, this relationship is captured in Figure~\ref{fig:inverted-U-expts}.%
\footnote{We consider the monopoly scenario for comparison only, without presenting any findings. We just assume that a monopolist chooses
the greedy algorithm, because it is easier to deploy in practice. Implicitly, users have no ``outside option": the service provided is an improvement over not having it (and therefore the monopolist is not incentivized to deploy better learning algorithms). This is plausible with free ad-supported platforms such as Yelp or Google.}

### 2020-07-15 journal paper/ec19paper/content/perf_in_iso.tex
Minor edits.

\begin{finding}
\textit{The mean reputation trajectories \asedit{and the instantaneous reward trajectories} are arranged as predicted by prior work:
    $\Thompson > \DynamicEpsGreedy > \DynamicGreedy$ for a sufficiently long time horizon.}
\end{finding}

>>>>

\begin{finding}
\textit{The mean reputation trajectories \asedit{and the instantaneous reward trajectories} are arranged as predicted by prior work:
    $\Thompson > \DynamicEpsGreedy > \DynamicGreedy$ for a sufficiently long time horizon $T$.}
\end{finding}
